Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-13 manuscript/manuscript.tex
pass through intro

Selection on both beneficial and deleterious alleles leads to 
perturbations in the allele frequency spectrum of neighboring, linked variation.
This phenomenon, known as linked selection, is a major force shaping patterns of
genetic variation along chromosomes \citep{Maynard_Smith1974-zr,Charlesworth1993-gb,Barton1998-aj,Gillespie2000-mh}.
As deleterious variation is clustered in functional portions of the genome,
we expect linked selection to reduce levels of diversity surrounding evolutionarily
conserved segments (e.g. genes, regulatory elements, etc.) 
and to shape large-scale spatial patterns in genetic diversity. 
This landscape of diversity is the product of a complex interplay between the
deleterious mutation rate, the local recombination rate, the strength of selection, and
the spatial distribution of conserved segments along a chromosome. 
Since genome-wide recombination maps and putatively conserved segments are available
for many species, there has been consistent effort to fit models of
linked selection to patterns of diversity to estimate population genetic
parameters such as the strength of selection and the deleterious mutation rate 
\citep{Hudson1995-pt,McVicker2009-ax},
the overall reduction in genetic variability due to the
selection on deleterious variants \citep{Elyashiv2016-vt,Murphy2022-sj},
and in some cases the rate of beneficial mutations \citep{corbett2015natural,Elyashiv2016-vt}.

Classic Background Selection (BGS) theory is the predominant model for the
reduction in linked neutral diversity used by statistical approaches to
quantify the impacts of negative selection. However, this model makes some
assumptions for mathematical tractability that could distort inferences about
the negative selection process. First, since fixation probabilities ultimately
depend on the product of the deleterious selection coefficient (s) and
population size (N), the efficacy of negative selection should depend on
demography. Unfortunately, accommodating demography into theoretic negative
selection models remains an open, difficult problem (XXX). Second, building off
classic models of mutation-selection balance (Crow 1970, Kimura and Maruyama
196), the Background Selection (XXX) model assumes that new mutations are
sufficiently deleterious that they are invariably driven to loss. Under this
assumption, the effect of selection is well-approximated by simply rescaling
the neutral coalescent by a reduction factor known as B (XXX). However, this is
only appropriate in a limited domain of parameters relevant to the negative
selection process (XXX). Because the BGS model cannot accommodate the
possibility of fixation for mutations with $Ns \ll 1$, diversity is incorrectly
predicted to fall to zero as the strength of selection diminishes. This weak
selection problem is related to the wider problem of modeling the substitution
rate of deleterious mutation (i.e. the “ratchet” rate). Third, the BGS model
assumes segments under purifying selection are sufficiently far apart that
there is no selective interference between them. This is a matter of degree,
since under sufficiently high mutation rates or with little recombination,
selective interference can occur with a segment.

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of the classic BGS model.
These models consider how polygenic fitness variation increases the rate of
stochastic allele frequency change as alleles become randomly linked to fitness
backgrounds. While these models can theoretically accommodate polygenic fitness
variation from any source as long as its rate of change is not too rapid, we
focus specifically on a deleterious-mutations-only model from Santiago and
Caballero (2016). This model is identical to the classic BGS model when
selection against deleterious mutations is strong, but it also correctly
predicts the reduction in diversity and substitutions rates of weakly
deleterious mutations by accurately modeling levels of genic fitness variation.
We extend Santiago and Caballero’s (hereafter the SC16) model so that it can be
used to model genome-wide patterns of diversity under negative selection
processes, and develop Python software to use this theoretic approach to fit
genome-wide models of negative selection. Using forward simulations, we show
this model leads to more accurate DFE estimates under weak selection. While
this new model accommodates the possibility that a given reduction B could be
due to weaker selection, in applying our composite-likelihood method to human
population genomic data, we confirm earlier findings that strong negative
selection at a small fraction of sites fits human data across populations best.
Finally, we show that while this model solves some limitations of the classic
BGS model, simulations reveal another mysterious defect: a poor fit around
$Ns=1$. We show that this can be partially corrected by incorporating the effects
of selective interference on genic variation, but the remaining discrepancy is
due to the build up of negative linkage disequilibrium that cannot be
adequately captured by this model. We find accounting for genic selective
interference between segments fits human population genomic data slightly
better, suggesting levels of negative selection in humans are sufficient to
perturb the selection at other sites.

>>>>

Selection on both beneficial and deleterious alleles leads to 
perturbations in the allele frequency spectrum of neighboring, linked variation.
This phenomenon, known as linked selection, is a major force shaping patterns of
genetic variation along chromosomes 
\parencite{Maynard_Smith1974-zr,Charlesworth1993-gb,Barton1998-aj,Gillespie2000-mh}.
As deleterious variation is clustered in functional portions of the genome,
we expect linked selection to reduce levels of diversity surrounding evolutionarily
conserved segments (e.g. genes, regulatory elements, etc.) 
and to shape large-scale spatial patterns in genetic diversity. 
This landscape of diversity is the product of a complex interplay between the
deleterious mutation rate, the local recombination rate, the strength of selection, and
the spatial distribution of conserved segments along a chromosome. 
Since genome-wide recombination maps and putatively conserved segments are available
for many species, there has been consistent effort to fit models of
linked selection to patterns of diversity to estimate population genetic
parameters such as the strength of selection and the deleterious mutation rate 
\parencite{Hudson1995-pt,McVicker2009-ax},
the overall reduction in genetic variability due to the
selection on deleterious variants \parencite{Elyashiv2016-vt,Murphy2022-sj},
and in some cases the rate of beneficial mutations \parencite{corbett2015natural,Elyashiv2016-vt}.

Prior attempts to model the reduction in linked diversity due to deleterious alleles
have largely relied on the now canonical Background Selection (BGS) model \parencite{Charlesworth1993-gb,Nordborg1996-nq}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions
may distort inferences about the negative selection process.
First, since fixation probabilities ultimately
depend on the product of the deleterious selection coefficient ($s$) and
population size ($N$), the efficacy of negative selection should depend on
demography. Unfortunately, accommodating demography into analytic models of negative
selection remains an open, difficult problem (XXX). Second, building off
classic models of mutation-selection balance \parencite{Crow1970-wj,Kimura1966-bk}
BGS assumes that new mutations are sufficiently deleterious
such that they are invariably driven to loss. 
Under this assumption, the effect of selection is well-approximated by simply rescaling
the coalescent effective size by a reduction factor, 
sometimes called $B$ such that $N_eB=N_{0}$ \parencite{Charlesworth2013-kl}.
However this rescaling is not appropriate across parts of parameter space
that are relevant to natural populations segregating deleterious variation
 \parencite[e.g.][]{good2014genetic,Santiago2016-mu}.
In particular, because the BGS model cannot accommodate the
possibility of fixation for mutations with $Ns \ll 1$, diversity is incorrectly
predicted to fall to zero as the strength of selection diminishes. This weak
selection problem is related to the wider problem of modeling the substitution
rate of deleterious mutation (i.e. the “ratchet” rate). Third, the BGS model
assumes segments under purifying selection are sufficiently far apart that
there is no selective interference between them. This is a matter of degree,
since under sufficiently high mutation rates or with little recombination,
selective interference can occur with a segment.

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of the classic BGS model 
\parencite{santiago1998effective,Santiago2016-mu}.
These models consider how polygenic fitness variation increases the rate of
stochastic allele frequency change as alleles become randomly linked to fitness
backgrounds. While these models can theoretically accommodate polygenic fitness
variation from any source as long as its rate of change is not too rapid, we
focus specifically on a deleterious-mutations-only model from Santiago and
Caballero (2016). 
This model is identical to BGS when selection against deleterious mutations is strong,
but it also correctly predicts the reduction in diversity and substitution rates of weakly
deleterious mutations by accurately modeling levels of genic fitness variation.
We extend Santiago and Caballero model (hereafter the SC16) so that it can be
used to model genome-wide patterns of diversity,
and develop Python software to use this theoretic approach to fit
genome-wide models of negative selection. Using forward simulations, we show
this model leads to more accurate DFE estimates under weak selection. 
Finally, we apply our composite-likelihood method to human population genomic data
and provide new estimates of the landscape of negative selection in humans.

%
%
% ADK- Vince I commented out the below because I don't like to walk through the results
% in an intro. Instead say what you did, and then the results are in the results section.
%
%While this new model accommodates the possibility that a given reduction B could be
%due to weaker selection, in applying our composite-likelihood method to human
%population genomic data, we confirm earlier findings that strong negative
%selection at a small fraction of sites fits human data across populations best.
%Finally, we show that while this model solves some limitations of the classic
%BGS model, simulations reveal another mysterious defect: a poor fit around
%$Ns=1$. We show that this can be partially corrected by incorporating the effects
%of selective interference on genic variation, but the remaining discrepancy is
%due to the build up of negative linkage disequilibrium that cannot be
%adequately captured by this model. We find accounting for genic selective
%interference between segments fits human population genomic data slightly
%better, suggesting levels of negative selection in humans are sufficient to
%perturb the selection at other sites.
